Top-down versus bottom-up cohesiveness in microbial community coalescence

### 2021-05-26 draft/draft_v3.tex
draft almost done

@@

To better understand the underlying mechanisms that govern
the emergence of ecological co-selection,
we developed a minimal model of community coalescence.
This model is comprised of two communities with two species each
as illustrated in \figref[a]{fig4}.
Within each community, the dominant species
($\mathrm{s}_1$ and $\mathrm{s}_3$ in the resident and invasive communities
respectively)
is able to utilize the primary resource that is supplied externally ($\mathrm{R}_1$).
As a result of $\mathrm{R}_1$ metabolism,
the dominants secrete a single byproduct
($\mathrm{R}_2$ and $\mathrm{R}_4$ respectively)
off which
the sub-dominants ($\mathrm{s}_2$ and $\mathrm{s}_4$ respectively) can feed.
Finally, these sub-dominants secrete an additional resource
($\mathrm{R}_3$ and $\mathrm{R}_5$ respectively)
that can again
be metabolized by the respective dominants.
This is the simplest structure where both top-down and bottom-up
cohesiveness can emerge, with their relative strengths being determined
by how effectively the species can utilize each other's metabolic byproducts.
Several parameters can modulate the strength and direction of community cohesiveness,
and even in this simple model there can be complex interactions between them.
For instance, a sub-dominant that is very efficient at metabolizing the dominant's
secretions will therefore be a strong producer of secondary byproducts,
in turn increasing the metabolic flow towards the dominant itself.
To keep both communities symmetrical and
mantain tractability of the model's behavior, we chose to make
the two sub-dominant species equally able consumers of the
byproduct of their respective dominants
(\methodsref{sim-min}).
We also made it so species secrete only one type of byproduct each,
i.e. the metabolic matrix is binary in this case
(see \hyperref[box1]{Box~1}).
The secretions of the resident and invasive communities are
non-overlapping:
\iffalse
(i.e. $\mathrm{s}_1$ and $\mathrm{s}_2$ secretions
differ from $\mathrm{s}_3$ and $\mathrm{s}_4$ secretions)
\fi
this ensures that coalescence always results in one of the two
communities taking over and completely excluding the other one,
which facilitates the interpretation of the model's outcomes.

With these constraints, the model is specified by four rates:
the resident dominant's uptake rate of the primary resource ($c_{11}$)
and of the secondary byproduct secreted by $\mathrm{s}_2$ ($c_{13}$),
as well as
the invasive dominant's uptake rate of the primary resource ($c_{31}$)
and of the secondary byproduct secreted by $\mathrm{s}_4$ ($c_{35}$).
Analogous to what we did for our initial
experiments and simulations of community coalescence,
we first stabilized each community separately
through serial growth-dilution cycles,
and then mixed them 1:1 and stabilized again until generational
equilibrium was achieved.
Equations \ref{eq:crm-n} and \ref{eq:crm-r} in \hyperref[box1]{Box~1}
describe the dynamics of this system between consecutive dilutions
with the corresponding choices for the parameter values
(\methodsref{sim-min}).
We ran simulations for both community-community competitions and
invasions from the invasive dominant alone.
We considered four different scenarios illustrated in \figref[b-e]{fig4}:
\textit{1)} both communities are cohesive
from the top-down but not from the bottom-up,
\textit{2)} the resident community is cohesive
both from the top-down and from the bottom-up
while
the resident community is cohesive
from the top-down only,
\textit{3)} the resident community is cohesive
from the top-down only
while
the invasive community is cohesive
both from the top-down and from the bottom-up,
and \textit{4)} both communities are cohesive
from the top-down and from the bottom-up.
The outcomes in each case are discussed below.

\paragraph*{Top-down resident versus top-down invasive.}
This is the simplest scenario.
When both communities are cohesive strictly from the top-down
($c_{13}=0$ and $c_{35}=0$),
coalescence outcomes are determined by which dominant is a better consumer
of the primary resource.
The most competitive dominant excludes the least competitive one
($\mathrm{s}_1$ excludes $\mathrm{s}_3$ if $c_{11}>c_{31}$ and vice-versa,
\figref[b]{fig4})
and,
because the cross-feeding networks of the invasive and resident
communities are non-overlapping,
it co-selects its sub-dominant exclusively.
The relative similarity between the invasive and the coalesced communities
($Q$, see \methodsref{metrics})
can only be 0 (if the resident dominant ``wins'' competition for
$\mathrm{R}_1$) or 1 (if the opposite happens).
The invasion success of the invasive dominant ($\mathrm{s}_3$) is
unaffected by the presence or absence of its sub-dominant ($\mathrm{s}_4$)
since it is not being cross-fed by it.
Thus, again $\mathrm{s}_3$ is only able to invade alone when it
outcompetes the resident dominant ($\mathrm{s}_1$) for the primary resource ($c_{11}>c_{31}$).

\paragraph*{Bidirectional resident versus top-down invasive.}
Blablabla...

\paragraph*{Top-down resident versus bidirectional invasive.}
Blablabla...

\paragraph*{Bidirectional resident versus bidirectional invasive.}
Blablabla...

\subsubsection*{Community hierarchy regulates the strength of bottom-up co-selction}

@@

@@

% references
\clearpage
\bibliographystyle{mystyle}
\bibliography{refs}

@@

\begin{figure}[!h]
\centering
\internallinenumbers
\includegraphics[scale=0.7,keepaspectratio]{figs/fig4_v2.pdf}
\caption{\textbf{A minimal model of community coalescence.}
\textbf{a.} Illustration of the model structure and parameters.
The primary resource ($\mathrm{R}_1$) is replenished after
each growth-dilution cycle (red arrows).
Solid arrows indicate resource consumption, dashed arrows represent
resource secretion.
\textbf{b-e.} Coalescence outcomes in the minimal model
under different relations of cohesiveness between the
resident and the invasive communities.}
\label{fig4}
\end{figure}

>>>>

\makeatother

@@

To better understand the underlying mechanisms that govern
the emergence of ecological co-selection,
we developed a minimal model of community coalescence.
This model is comprised of two communities with two species each
as illustrated in \figref[a]{fig4}.
Within each community, the dominant species
($\mathrm{s}_1$ and $\mathrm{s}_3$ in the resident and invasive communities
respectively)
is able to utilize the primary resource that is supplied externally ($\mathrm{R}_1$).
As a result of $\mathrm{R}_1$ metabolism,
the dominants secrete a single byproduct
($\mathrm{R}_2$ and $\mathrm{R}_4$ respectively)
off which
the sub-dominants ($\mathrm{s}_2$ and $\mathrm{s}_4$ respectively) can feed.
Finally, these sub-dominants secrete an additional resource
($\mathrm{R}_3$ and $\mathrm{R}_5$ respectively)
that can again
be metabolized by the corresponding dominants.
This is the simplest structure where both top-down and bottom-up
cohesiveness can emerge, with their relative strengths being determined
by how effectively the species can utilize each other's metabolic byproducts.
Several parameters can modulate the intensity and direction of community cohesiveness,
and even in this simple model there can be complex interactions between them.
For instance, a sub-dominant that is very efficient at metabolizing the dominant's
secretions will therefore be a strong producer of secondary byproducts,
in turn increasing the metabolic flow towards the dominant itself.
To keep both communities symmetrical and
mantain tractability of the model's behavior, we chose to make
the two sub-dominant species equally able consumers of the
byproduct of their respective dominants
(\methodsref{sim-min}).
We also made it so species secrete only one type of byproduct each,
i.e. the metabolic matrix is binary in this case
(see \hyperref[box1]{Box~1}).
The secretions of the resident and invasive communities are
non-overlapping:
\iffalse
(i.e. $\mathrm{s}_1$ and $\mathrm{s}_2$ secretions
differ from $\mathrm{s}_3$ and $\mathrm{s}_4$ secretions)
\fi
this ensures that coalescence always results in one of the two
communities taking over and completely excluding the other one,
which facilitates the interpretation of the model's outcomes.
This means that
the relative similarity between the invasive and the coalesced communities
($Q$, see \methodsref{metrics})
can only be 0 (if the resident community dominates in coalescence)
or 1 (if the opposite happens).

With these constraints, the model is specified by four rates:
the resident dominant's uptake rate of the primary resource ($c_{11}$)
and of the secondary byproduct secreted by $\mathrm{s}_2$ ($c_{13}$),
as well as
the invasive dominant's uptake rate of the primary resource ($c_{31}$)
and of the secondary byproduct secreted by $\mathrm{s}_4$ ($c_{35}$).
Analogous to what we did for our initial
experiments and simulations of community coalescence,
we first stabilized each minimal community separately
through serial growth-dilution cycles,
and then mixed them 1:1 and stabilized again until generational
equilibrium was achieved.
Equations \ref{eq:crm-n} and \ref{eq:crm-r} in \hyperref[box1]{Box~1}
describe the dynamics of this system between consecutive dilutions
with the corresponding choices for the parameter values
(\methodsref{sim-min}).
We ran simulations for both community-community competitions and
invasions from the dominant alone.
We considered four different scenarios illustrated in \figref[b-e]{fig4}:
\textit{1)} both communities are cohesive
from the top-down but not from the bottom-up,
\textit{2)} the resident community is cohesive
both from the top-down and from the bottom-up
while
the resident community is cohesive
from the top-down only,
\textit{3)} the resident community is cohesive
from the top-down only
while
the invasive community is cohesive
both from the top-down and from the bottom-up,
and \textit{4)} both communities are cohesive
from the top-down and from the bottom-up.
The outcomes in each case are discussed below.

\textbf{Top-down resident versus top-down invasive.}
This is the simplest scenario.
When both communities are cohesive strictly from the top-down
($c_{13}=0$ and $c_{35}=0$),
coalescence outcomes are determined by which dominant is a better consumer
of the primary resource.
The most competitive dominant excludes the least competitive one
($\mathrm{s}_1$ excludes $\mathrm{s}_3$ if $c_{11}>c_{31}$ and vice-versa,
\figref[b]{fig4})
and,
because the cross-feeding networks of the invasive and resident
communities are non-overlapping,
it co-selects its sub-dominant exclusively.
The invasion success of the invasive dominant ($\mathrm{s}_3$) is
unaffected by the presence or absence of its sub-dominant ($\mathrm{s}_4$)
since it is not being cross-fed by it.
Thus, again $\mathrm{s}_3$ is only able to invade alone when it
outcompetes $\mathrm{s}_1$ for the primary resource ($c_{11}>c_{31}$).

\textbf{Bidirectional resident versus top-down invasive.}
When
the resident community is cohesive both top-down and bottom-up ($c_{13}>0$)
and the invasive community is only cohesive from the top-down ($c_{35}=0$),
coalescence outcomes
are determined by the strength of bottom-up cohesiveness in the resident community,
i.e. the magnitude of $c_{13}$ in the minimal model
(\figref[c]{fig4}).
If $c_{13}$ is small (weak bottom-up cohesiveness), the system behaves as in the previous scenario,
virtually as if both communities were only cohesive from the top-down.
However, as $c_{13}$ increases (strong bottom-up cohesiveness) the resident community becomes more
resistant to invasion.
For large values of $c_{13}$, the invasive dominant being more competitive than the resident
dominant for the primary resource does not ensure invasion success.
Since the latter can also be cross-fed by its sub-dominant (through $\mathrm{R}_3$),
it does not rely on the primary resource only ($\mathrm{R}_1$) to outcompete the invader.
The same is true when the invasive dominant invades alone.

\textbf{Top-down resident versus bidirectional invasive.}
More outcomes become possible if the resident community is cohesive from the top-down only ($c_{13}=0$)
and the invasive one is cohesive in both directions ($c_{35}>0$).
Like in the first scenario we considered, the invasive dominant will only be successful by itself if it is a better
competitor for $\mathrm{R}_1$ than the resident dominant ($c_{31}>c_{11}$).
However, if the invasive sub-dominant is also present, the invasion can be successful even if $c_{31}<c_{11}$
as long as the cross-feeding from $\mathrm{s}_4$ (invasive sub-dominant) to $\mathrm{s}_3$ (invasive dominant)
is strong enough, i.e. if $c_{35}$ sufficiently large
(\figref[d]{fig4}).
In this scenario, we can see cases where the invasive dominant cannot invade in isolation but
the invasive community still dominates during coalescence,
similar to what we observed in our experiments and initial simulations.

\textbf{Bidirectional resident versus bidirectional invasive.}
If both communities are cohesive from the top-down and from the bottom-up ($c_{13}>0$ and $c_{35}>0$),
coalescence outcomes are generically determined by the relations of strength between
all forms of cohesiveness.
In limit cases we can recover scenarios considered previously, e.g. if $c_{13} >> c_{35}$ the system
behaves as if the resident community was cohesive from the bottom-up but the invasive community was not.
Situations where the invasive dominant can invade when accompanied by its sub-dominant despite
being unable to do so by itself are again possible
if the invasive community is more strongly cohesive from the bottom-up than the resident community
(\figref[e]{fig4}).

This minimal coalescence model shows that
bottom-up co-selection of a dominant that is unable to invade by itself
is only possible if said dominant is strongly cross-fed by its cohort.
The model also demonstrates that this behavior can only happen
if the byproducts of the cohorts are to some extent non-overlapping
and preferentially utilized by the respective dominants.
Otherwise, if both species can utilize all secretions, bottom-up
co-selection becomes unspecific with respect to the dominant identity.

\subsubsection*{Community hierarchy regulates the strength of bottom-up co-selection}

@@

Our minimal model is set within the same MicroCRM framework that we used
for the previous simulations.
As described in the main text, the model contains two communities
of two species each
($\mathrm{s}_1$ to $\mathrm{s}_4$),
with five resources in total, out of which the first one ($\mathrm{R}_1$)
is replenished externally at the beginning of each growth cycle
and the rest correspond to the species' metabolic byproducts.
Each species secretes a unique byproduct, meaning that the metabolic
matrix $\mathbf{D}$ is binary in this case.
The specific structure of $\mathbf{D}$ is displayed below
--because it is a 3-dimensional matrix in our framework,
we have ``sliced'' it into the five 2-dimensional matrices corresponding to
our five species.

\begin{equation*}
\mathbf{D}_1 = \; \; \; \; \; \begin{matrix}
 &

\begin{matrix}
\text{\scriptsize $\mathrm{R}_1$} \!\!\! & 
\text{\scriptsize $\mathrm{R}_2$} \!\!\! & 
\text{\scriptsize $\mathrm{R}_3$} \!\!\! & 
\text{\scriptsize $\mathrm{R}_4$} \!\!\! & 
\text{\scriptsize $\mathrm{R}_5$}
\end{matrix}

\\

\begin{matrix*}[l]
\text{\scriptsize $\mathrm{R}_1$} \!\!\! \\
\text{\scriptsize $\mathrm{R}_2$} \!\!\! \\
\text{\scriptsize $\mathrm{R}_3$} \!\!\! \\
\text{\scriptsize $\mathrm{R}_4$} \!\!\! \\
\text{\scriptsize $\mathrm{R}_5$} \!\!\!
\end{matrix*}

&

\begin{pmatrix}
0 & * & 0 & * & * \\
1 & * & 0 & * & * \\
0 & * & 1 & * & * \\
0 & * & 0 & * & * \\
0 & * & 0 & * & *
\end{pmatrix}

\end{matrix}
\end{equation*}

\bigskip

\begin{equation*}
\mathbf{D}_2 = \; \; \; \; \; \begin{matrix}
 &

\begin{matrix}
\text{\scriptsize $\mathrm{R}_1$} \!\!\! & 
\text{\scriptsize $\mathrm{R}_2$} \!\!\! & 
\text{\scriptsize $\mathrm{R}_3$} \!\!\! & 
\text{\scriptsize $\mathrm{R}_4$} \!\!\! & 
\text{\scriptsize $\mathrm{R}_5$}
\end{matrix}

\\

\begin{matrix*}[l]
\text{\scriptsize $\mathrm{R}_1$} \!\!\! \\
\text{\scriptsize $\mathrm{R}_2$} \!\!\! \\
\text{\scriptsize $\mathrm{R}_3$} \!\!\! \\
\text{\scriptsize $\mathrm{R}_4$} \!\!\! \\
\text{\scriptsize $\mathrm{R}_5$} \!\!\!
\end{matrix*}

&

\begin{pmatrix}
* & 0 & * & * & * \\
* & 0 & * & * & * \\
* & 1 & * & * & * \\
* & 0 & * & * & * \\
* & 0 & * & * & *
\end{pmatrix}

\end{matrix}
\end{equation*}

\bigskip

\begin{equation*}
\mathbf{D}_3 = \; \; \; \; \; \begin{matrix}
 &

\begin{matrix}
\text{\scriptsize $\mathrm{R}_1$} \!\!\! & 
\text{\scriptsize $\mathrm{R}_2$} \!\!\! & 
\text{\scriptsize $\mathrm{R}_3$} \!\!\! & 
\text{\scriptsize $\mathrm{R}_4$} \!\!\! & 
\text{\scriptsize $\mathrm{R}_5$}
\end{matrix}

\\

\begin{matrix*}[l]
\text{\scriptsize $\mathrm{R}_1$} \!\!\! \\
\text{\scriptsize $\mathrm{R}_2$} \!\!\! \\
\text{\scriptsize $\mathrm{R}_3$} \!\!\! \\
\text{\scriptsize $\mathrm{R}_4$} \!\!\! \\
\text{\scriptsize $\mathrm{R}_5$} \!\!\!
\end{matrix*}

&

\begin{pmatrix}
0 & * & * & * & 0 \\
0 & * & * & * & 0 \\
0 & * & * & * & 0 \\
1 & * & * & * & 0 \\
0 & * & * & * & 1
\end{pmatrix}

\end{matrix}
\end{equation*}

\bigskip

\begin{equation*}
\mathbf{D}_4 = \; \; \; \; \; \begin{matrix}
 &

\begin{matrix}
\text{\scriptsize $\mathrm{R}_1$} \!\!\! & 
\text{\scriptsize $\mathrm{R}_2$} \!\!\! & 
\text{\scriptsize $\mathrm{R}_3$} \!\!\! & 
\text{\scriptsize $\mathrm{R}_4$} \!\!\! & 
\text{\scriptsize $\mathrm{R}_5$}
\end{matrix}

\\

\begin{matrix*}[l]
\text{\scriptsize $\mathrm{R}_1$} \!\!\! \\
\text{\scriptsize $\mathrm{R}_2$} \!\!\! \\
\text{\scriptsize $\mathrm{R}_3$} \!\!\! \\
\text{\scriptsize $\mathrm{R}_4$} \!\!\! \\
\text{\scriptsize $\mathrm{R}_5$} \!\!\!
\end{matrix*}

&

\begin{pmatrix}
* & * & * & 0 & * \\
* & * & * & 0 & * \\
* & * & * & 0 & * \\
* & * & * & 0 & * \\
* & * & * & 1 & *
\end{pmatrix}

\end{matrix}
\end{equation*}
%
Asterisks indicate values of $D_{i\beta\alpha}$ that are irrelevant because
species $i$ cannot utilize resource $\alpha$ (and so the metabolic flux from
$\alpha$ to $\beta$ corresponding to that species will always be zero
regardless of the value of $D_{i\alpha\beta}$).
The consumer preference matrix $\mathbf{c}$ takes the following form:

\begin{equation*}
\mathbf{c} = \; \; \; \; \; \begin{matrix}
 &

\begin{matrix}
\text{\scriptsize $\mathrm{R}_1$} \;\;\; & 
\text{\scriptsize $\mathrm{R}_2$} \;\;\; & 
\text{\scriptsize $\mathrm{R}_3$} \;\;\; & 
\text{\scriptsize $\mathrm{R}_4$} \;\;\; & 
\text{\scriptsize $\mathrm{R}_5$}
\end{matrix}

\\

\begin{matrix*}[l]
\text{\scriptsize $\mathrm{s}_1$} \!\!\! \\
\text{\scriptsize $\mathrm{s}_2$} \!\!\! \\
\text{\scriptsize $\mathrm{s}_3$} \!\!\! \\
\text{\scriptsize $\mathrm{s}_4$} \!\!\!
\end{matrix*}

&

\begin{pmatrix}
c_{11} & 0 & c_{13} & 0 & 0 \\
0 & 100 & 0 & 0 & 0 \\
c_{31} & 0 & 0 & 0 & c_{35} \\
0 & 0 & 0 & 100 & 0
\end{pmatrix}

\end{matrix}
\end{equation*}
%
Where we made the sub-dominants equally strong consumers of their dominants' secretions
($c_{22} = c_{44}$ = 100),
and we varied all other uptake rates depending on the scenario we were considering (see main text).
Whenever we were interested in the ratio between two rates (e.g. $c_{35}/c_{13}$ in \figref[e]{fig4})
we gave the one in the denominator a fixed value of 1 and let one in the numerator range
within the specified limits.

@@

% references
\clearpage
\bibliographystyle{mystyle}
\bibliography{refs}
\clearpage

@@

\begin{figure}[!h]
\centering
\internallinenumbers
\includegraphics[scale=0.7,keepaspectratio]{figs/fig4_v3.pdf}
\caption{\textbf{A minimal model of community coalescence.}
\textbf{a.} Illustration of the model structure and parameters.
The primary resource ($\mathrm{R}_1$) is replenished after
each growth-dilution cycle (red arrows).
Solid arrows indicate resource consumption, dashed arrows represent
resource secretion.
\textbf{b-e.} Coalescence outcomes in the minimal model
under different relations of cohesiveness between the
resident and the invasive communities.}
\label{fig4}
\end{figure}
